Protected Areas Practical

### 2024-11-25 tutorial/MSc_PA_practical_answers.Rmd
stash went weird

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical.

@@

- directly estimated: we can do an economic evaluation about economic losses of converting land to  protected land. This can be through the cost of purchasing land, costing the staff needed to patrol, the revenue from tourism, the loss of opportunities (loss of mining land, grazing land, agricultural land, and even homes).
- indirectly estimated: we can estimate what land is most costly. For example, the human footprint index is often used to see where people are using land. We then make the assumption that areas that people use are most costly and created the greatest conflict, and should be avoided as a place to put a protected area.

Here, it's an excecise, so we use the Human footprint layer, which is downloaded from this paper:
https://www.nature.com/articles/s41597-022-01284-8#Sec12

@@

# Spatial extent

@@

If this is your first time using the prioritizr R package, please install the HiGHS solver using the following R code. Although the HiGHS solver is relatively fast and easy to install. If later you really get into protected area planning, then use your Queen Mary VPN to install the Gurobi software suite and the gurobi R package for best performance (see the [Gurobi Installation Guide for details](https://prioritizr.net/articles/gurobi_installation_guide.html)).

@@

 **I want to protect as many species as possible within my budget by setting cells aside for protection, aiming for 40% of the range of each species**

@@

- **Budget**: the total cost of the prioritization will represent a 22% of the total land value in the study area. Given this budget, we want the prioritization to increase feature representation, as much as possible, 
- **Features**: each feature (i.e. species) would, ideally, have 30% of its distribution covered by the prioritization. In this scenario, we can either purchase all of the land inside a given planning unit, or none of the land inside a given planning unit. 
- **Planning Units**: These are sites that we are considering to protect or not. We have the option to protect or not each site. In this case, each site represents one pixel of the raster

@@

## Extract the objective

```{r}
print(attr(s1, "objective"))
```
## Get the time it took to solve

@@

@@

# Exercise 3

Modify the edge factor to 0.2 and the penalty to 0.7 and see what happens

```{r}
pz <-
  p2 %>%
  add_boundary_penalties(penalty = 0.7, edge_factor = 0.2)

# solve the problem
sz <- solve(pz)

# plot the solution
plot(sz, main = "Solution", axes = FALSE)

# calculate target coverage for the solution
pz_target_coverage <- eval_target_coverage_summary(pz, sz)
print(pz_target_coverage)

# check percentage of the features that have their target met given the solution
print(mean(pz_target_coverage$met) * 100)
```

Reducing the edge factor makes it even blockier!

@@

# Exercise 4: Lock out land from conservation

@@

# Exercise 5: Implementing Marxan

@@

# changed the plot colours because why not
# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s1, main = "maximise biodiversity", axes = FALSE)
plot(sy, main = "minimise cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot

@@

>>>>

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical. The software we will use to do this is `prioritizr`.

@@

- no cost: cost is equal across all planning units, which means the prioritisation will only rely on the biodiversity data. However, a protected area plan that doesn't account for cost is not likely to be very helpful.
- directly estimated: we can do an economic evaluation about economic losses of converting land to  protected land. This can be through the cost of purchasing land, costing the staff needed to patrol, the revenue from tourism, the loss of opportunities (loss of mining land, grazing land, agricultural land, and even homes).
- indirectly estimated: we can estimate what land is most costly. For example, the human footprint index is often used to see where people are using land. We then make the assumption that areas that people use are most costly and create the greatest conflict, and should be avoided as a place to put a protected area.

Here, it's an execise, so we use the Human footprint layer, which is downloaded from this paper:
https://www.nature.com/articles/s41597-022-01284-8#Sec12

@@

# Spatial extent and projection

@@

If this is your first time using the prioritizr R package, please install the HiGHS solver using the following R code. Although the HiGHS solver is relatively fast and easy to install. If later you really get into protected area planning, then use your Queen Mary VPN to install the Gurobi software suite and the Gurobi R package for best performance (see the [Gurobi Installation Guide for details](https://prioritizr.net/articles/gurobi_installation_guide.html)).

@@

 **I want to protect as many species as possible within my budget by setting cells aside for protection, aiming for 30% of the range of each species**

@@

- **Budget**: the total cost of the prioritization will represent a 30% of the total land value in the study area. Given this budget, we want the prioritization to increase feature representation, as much as possible, 
- **Features**: each feature (i.e. species) would, ideally, have 30% of its distribution covered by the prioritization. In this scenario, we can either purchase all of the land inside a given planning unit, or none of the land inside a given planning unit. 
- **Planning Units**: These are sites that we are considering to protect or not. We have the option to protect or not each site. In this case, each site represents one pixel of the raster

@@

## Get the time it took to solve

@@

Hint: replace the value in `add_relative_targets()`

@@

@@

# Exercise 3: Lock out land from conservation

@@

# Exercise 4: Implementing Marxan

@@

# changed the plot colours because why not
# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s2, main = "maximise biodiversity", axes = FALSE)
plot(sy, main = "minimise cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot

@@

# Exercise 5: Ignore cost

So far we have used cost as the human footprint layer, but what if we want to just ignore cost and look only at the biodiversity? We are going to create a new cost layer called `cost_null` where cost is set to 1 on land, and NA everywhere else. We will then  replacee `cost` with `cost_null` in` problem(cost, features = species)`. How does it compare to the layer with cost?

Hint: replace all `cost` with `cost_null` in `budget` and `pr`

```{r}
cost_null = ifel(!is.na(cost),1,NA)

# calculate budget
budget <- terra::global(cost_null, "sum", na.rm = TRUE)[[1]] * 0.3# replace cost with cost_null here

# create problem
pr <-
  problem(cost_null, features = species) %>% # replacee cost with cost_null in problem(cost, features = species) %>% 
  add_min_shortfall_objective(budget) %>%
  add_relative_targets(0.3) %>%
  add_binary_decisions() %>%
  add_default_solver(gap = 0.1, verbose = FALSE)

sr <- solve(pr)

# changed the plot colours because why not
# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s1, main = "with cost", axes = FALSE)
plot(sr, main = "without cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot

### 2024-11-25 tutorial/MSc_PA_practical_exercise.Rmd
stash went weird

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical.

@@

# Exercise 3

Modify the edge factor to 0.2 and the penalty to 0.7 and see what happens

```
pz <-
  p2 %>%
  add_boundary_penalties(penalty = 0.003, edge_factor = 0.5) # change the values here

# solve the problem
sz <- solve(pz)

# plot the solution
plot(sz, main = "Solution", axes = FALSE)

# calculate target coverage for the solution
pz_target_coverage <- eval_target_coverage_summary(pz, sz)
print(pz_target_coverage)

# check percentage of the features that have their target met given the solution
print(mean(pz_target_coverage$met) * 100)
```

Reducing the edge factor makes it even blockier!

@@

# changed the plot colours because why not
# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s1, main = "maximise biodiversity", axes = FALSE)
plot(sy, main = "minimise cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot
```

>>>>

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical.The software we will use to do this is `prioritizr`.

@@

@@

# changed the plot colours because why not
# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s1, main = "maximise biodiversity", axes = FALSE)
plot(sy, main = "minimise cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot
```
# Exercise 5: Ignore cost

So far we have used cost as the human footprint layer, but what if we want to just ignore cost and look only at the biodiversity? We are going to create a new cost layer called `cost_null` where cost is set to 1 on land, and NA everywhere else. We will then  replacee `cost` with `cost_null` in` problem(cost, features = species)`. How does it compare to the layer with cost?

Hint: replace all `cost` with `cost_null` in `budget` and `pr`

```
cost_null = ifel(!is.na(cost),1,NA)

# calculate budget
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3 # replace cost with cost_null here

# create problem
pr <-
  problem(cost, features = species) %>% # replacee cost with cost_null here
  add_min_shortfall_objective(budget) %>%
  add_relative_targets(0.3) %>%
  add_binary_decisions() %>%
  add_default_solver(gap = 0.1, verbose = FALSE)

sr <- solve(pr)

# plot the solution
par(mfrow=c(1,2))# split into 2 plots side by side
plot(s1, main = "with cost", axes = FALSE)
plot(sr, main = "without cost", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot
